This draft supports the recoverable-resilience framing. The empirical data reveal rainfall-aligned mobility disruptions, nonuniform speed-deficit burden, OD functional dependence, and heterogeneous natural recovery. The optimization model then estimates the counterfactual portion of this loss that is recoverable under a standardized management regime.

The learning analysis extracts linked laws from the optimizer-derived recovery-value field. First, local marginal recovery value is activated when persistent future deficit overlaps with OD exposure and feasible intervention efficiency; one-factor rules based on highest deficit, highest exposure, or highest static bottleneck alone produce systematic false priorities. Second, finite-budget allocation must be residual: after each deployment, the remaining value depends on the current unrecovered loss, remaining budget, remaining time, and diminishing returns. Third, budget leverage is scale-dependent: absolute law-guided gain increases with budget, but relative and per-cost leverage decline in both the fine budget sweep and representative budget-specific LP closures. At the event level, decision-criticality is governed by the top tail of marginal recovery values rather than disruption magnitude alone.

The current results should be read as regime-conditional evidence, not as a final universal law. They also remain primarily hindsight counterfactual results, although early-window tests show partial predictability of decision-critical events. The factorized surrogate, event-regime validation, within-city chronological holdout, neural leakage audit, parameter-deconfounded tests, first-order and representative full-LP parameter ensembles, OD graph alignment ablation, and OD message-passing test suggest that the current law is low-dimensional but genuinely structural: it depends on the alignment between persistent loss, feasible intervention channels, and observed OD exposure, while explicit one-hop or two-hop OD messages add little once that alignment is represented. The next strongest extensions are wider scenario-specific LP closure, broader finite-budget intervention-parameter ensembles, zone-level event-footprint mapping, operational nowcasting, clean multi-year temporal validation within the same cities, and road-topology or graph-neural baselines that test whether the same symbolic law persists under richer spatial representations.
