The analysis distinguishes what the data can identify directly from what requires counterfactual modeling. Descriptive data mining reveals rainfall timing, speed-deficit abnormality, endogenous recovery, spatial concentration, and OD functional dependence. It cannot by itself identify the additional loss that would have been reduced by unobserved interventions. The optimization model supplies this counterfactual probe, and the learning analysis then compresses the optimizer-derived value field into interpretable laws.

The resulting laws are conditional rather than universal constants. They are learned under a standardized recovery regime: fixed intervention primitives, effectiveness assumptions, costs, deployment caps, response delays, budget intensity, and piecewise-linear diminishing returns. Within that regime, the findings are stable: first-order marginal value follows an activated exposure--horizon--efficiency form, while complete finite-budget allocation requires residual re-scoring to avoid saturation and overlapping effects. The non-obvious action-law analysis further shows that high deficit, high exposure, and high static bottleneck scores are not sufficient recovery rules; static structure acts as latent leverage only when it overlaps with persistent future loss, relevant OD exposure, feasible timing, and efficient intervention. The representative scenario-optimum validation further shows that this residual law remains close to non-base LP optima when budget and delay change, and the parameter-ensemble LP closures show the same pattern under eta, cost, delay, and channel-favored management regimes. The fine budget sweep adds a useful negative result: over budget scales 0.10--3.00, absolute decision leverage does not show an interior peak, while relative and per-budget leverage decline. Representative budget-specific LP closures confirm the same shape against re-solved optima, although they still cover one event per city rather than all 105 events.

The surrogate evidence also argues for parsimony. In the factorized leave-one-city tests, the largest decision-centered gains come from adding OD exposure and time-feasibility activation, while unrestricted high-dimensional interaction terms do not improve the full model. The event-regime tests point in the same direction: the factorized law remains close to the full additive surrogate when entire rainfall, impact, duration, loss, and time-of-day regimes are held out. The within-city chronological holdout gives a complementary temporal robustness check: holding out early, middle, late, or half-period event groups within each city still leaves the compact law with high top-tail capture, although the full additive surrogate retains a small accuracy margin. A neural-surrogate audit adds a guardrail against over-reading black-box fit: lightweight MLPs do not beat ridge laws under leave-city top-tail evaluation, and token-random identity features can inflate token-level correlation. Training-objective ablation adds a related boundary: top-tail capture and regret are the right evaluation metrics, but the compact law does not require a specialized ranking-loss objective in the current data. A parameter-deconfounded test adds a second boundary: action mechanics alone explain only about half of top-tail value, and a factorized law without calibrated \(\eta/c\) nearly matches the full factorized law, so the structure result is not simply a cost-effectiveness artifact. A first-order parameter-ensemble test extends this boundary: the parameter-light law transfers across eta, cost, delay, and channel-favored perturbations, while a base-trained surrogate with absolute \(\eta/c\) is more scale-sensitive. The OD graph ablation adds a sharper structural result: preserving city-level graph-feature distributions while shuffling their unit alignment removes much of the top-tail recovery-value signal. The OD message-passing surrogate refines this boundary: one-hop and two-hop OD neighborhood summaries are informative on their own, but add almost nothing once scalar OD exposure and scarcity are present. Thus city structure matters through actual OD spatial alignment, not only through aggregate exposure concentration, while the present action-value law remains low-dimensional enough to be written as an interpretable activation rule.

The main empirical message is therefore not that the largest disruptions are always the most recoverable. In the current sample, event-level decision-criticality is more closely tied to marginal-value top-tail concentration than to no-intervention loss magnitude. The severity-decoupling audit gives concrete counterexamples: some high-loss or high-rain events have low decision-criticality, while moderate-loss events can be highly decision-critical when value concentrates in the action top tail. This supports a decision-centered definition of recoverable resilience: events are most valuable for management when addressable loss is concentrated on a small set of exposed, intervention-sensitive locations.

The early-window analysis clarifies the operational boundary. Decision-criticality has early signals under leave-city validation, especially through static city structure and abnormal speed in the first hours, but recoverable fraction remains more dependent on the full observed trajectory. Therefore the present study should not claim full online dispatch capability. It establishes hindsight counterfactual recoverability and shows that some of its event-level signals may be partially visible early.

Several boundaries remain. The current sample contains seven full-zone cities with usable speed-event overlap. Scenario-specific LP closure is representative rather than exhaustive: V31 improves the non-base closure set to 26 of 28 optimal rows, but two large New York budget-only cases remain unresolved. Intervention effectiveness and cost parameters are not causally identified from observed intervention records, and the full-LP parameter-ensemble closure is still representative rather than exhaustive over all events and parameter scenarios. The event-regime, within-city chronological, and neural leakage audits are stronger than random token splits, but a clean calendar leave-time-period-out test is not yet available because years are strongly confounded with cities in the current sample. The event-level top-tail result is also partly city-structural under the current calibration: V32 shows zero within-city variation in top-5 percent value share for five of seven cities, because event deficits are still spatialized through a city-level OD vulnerability template. Without zone-level rainfall and speed footprints, within-city spatial variation across events is only partially resolved. Finally, the present graph result is an OD-dependency alignment ablation, not a full road-adjacency or graph-neural closure. A richer graph surrogate would strengthen the spatial representation test in a later version, but the current symbolic law is already reproducible from action-token evidence.
