The optimization model allocates high-level recovery primitives across space and time to reduce cumulative access-weighted functional loss. Durable restoration reduces underlying deficit, temporary capacity reduces locally experienced deficit, and substitution or control reduces origin-level experienced loss.

Let $b_{i,t}$ denote underlying functional deficit, $d_{i,t}$ local experienced deficit, and $\ell_{i,t}$ access-weighted functional loss. A functional dependence matrix $Q_t$ maps destination deficits to origin-level loss. The model minimizes $\sum_t \sum_i p_i \ell_{i,t}$ subject to recovery dynamics, resource budgets, response delays, and bounded functional states.

In the current computational implementation, the continuous model is solved as a linear program using Gurobi. The no-additional-intervention baseline fixes all recovery allocations to zero, while the managed-recovery counterfactual optimizes the three intervention primitives under period and total budget constraints. To avoid unrealistic concentration of all resources into a single high-leverage primitive, the implementation adds primitive-specific continuous deployment caps and concave piecewise-linear diminishing returns. These refinements preserve linearity because deployment is split into continuous segments with decreasing marginal effectiveness.

First-pass empirical calibration uses OD demand to estimate exposure weights and the dependence matrix, speed-deficit mining outputs to initialize functional deficit, rainfall-speed alignment to define short disturbance profiles, and observed recovery proxies to set endogenous recovery rates. Intervention effectiveness, deployment costs, budgets, and delays remain scenario parameters and are therefore evaluated through sensitivity sweeps rather than interpreted as observed causal effects.
